Tidy data structure to support exploration and modeling of temporal-context data

### 2018-08-08 main.Rmd
updated some thoughts

In the literature, time series and panel (longitudinal) data are common terms referred to temporal-context data, depending on the research fields. Researchers who are concerned with modelling large $T$ and small $N$ would name as "time series" (serial correlation); those who are interested in modelling small $T$ and large $N$ as "panel data" (asymptotic). The data format is two-dimensional array, but different modelling focuses lend the data input to different representations. A matrix is used to represent multivariate time series where each row represents observations measured at a time point and each column represents a series ("wide" form). This matrix representation requires homogeneity (that is, all the columns must be of same type.), time indices implicitly inferred as attributes or meta-information, series of same time length, and explicit missingness. By contrast, panel data are organised in rectangular form of heterogeneous column types where multiple study subjects are stacked and repeated for its time indices in a single column ("long" form), due to commonly arisen unbalanced panels. This specification requires explicitly declared panel variable and index, which has been implemented in Stata's time series module and R package **plm** [@R-plm]. Evidently, this data organisation appears more flexible than matrix in supporting explicit time index, multiple subjects of different time lengths, and implicit missing values. less defaults to be made. [R]

@@

The choice of representation of temporal-context data is made from a data-centric perspective, which is taken in the light of the operations that are to be performed on the data. This data abstraction reflects most common problems of incoming data and transformation in reality. Firstly, a data set must be structured in a "tidy" rectangular layout each row has the same schema and each field has a single value. Secondly, declaring the data set to contain temporal observations is determined by an "index" that represents time and a "key" that uniquely identifies each unit that measurements take place on over time. The "key" serves a similar purpose as the panel variable in the Stata's **tsset** command to define the units or subjects, but it is expanded to include multiple variables rather than a single one. A syntax is introduced to express a key consisting of nested and crossed factors. The composition of index and key uniquely defines each observation in a data table, which is equivalent to a primary key [@codd_relational_1970] in a relational database. Table \@ref(tab:tb-sub) shows a "tidy" temporal data example using a subset of tuberculosis cases estimated by @tb-data. It contains 12 observations and 5 variables arranged in a "long" tabular form. Each observation uniquely records the number of people, who are diagnosed tuberculosis for each gender at three selected countries in the years of 2011 and 2012, where column `year` is declared as the index variable, columns `country`, `continent` and `gender` as key. To be more specific, the key is passed via `id(country | continent, gender)` or `id(gender, country | continent)`: a geographic hierarchy of country nested within continent, crossed with a demographic factor. Detailed discussions about index and key specifications will be given in Section \@ref(sec:index) and \@ref(sec:key). Although column `count` is the only measure in this case, it is sufficiently flexible to hold other measured variables, for example, adding the corresponding population size (if known) in order to calibrate the count later.

Given the nature of temporal ordering, a temporal data set must be sorted by time index as a result. If a key is explicitly declared, the key will be sorted first depending on the ordering of factor levels and the ordering of variables, and followed by arranging time in ascending order. The tuberculosis example shows the key is sorted in alphabetic order, followed by ascending years by default, as is Table \@ref(tab:tb-sub). Since time-based operations, for example leading, lagging, or differencing series, require a vector in its time order, the arranged data format assures its validity for these operations without returning unexpected results. There are other possible arrangements in the key, but the baseline is the index (i.e. `year`) ascends within each key element. If any operations on the tidy temporal data results a deviation from this baseline, it starts issuing a warning to the users.

@@

The "key" identifies units or subjects that are observed over time in a data table, which are typically known in advance by users. The absence of key only occurs to a univariate case where the key is implicit. Multiple units must associate with an explicit "key" to allow for identification. The key is not constrained to a single column, but is comprised of multiple columns. When involving multiple variables, one need to distinguish nesting and crossing data variables. "Nesting" refers to a scenario where a variable is nested within another when each category of the former variable co-occurs with only one category of the latter (for example, `country` nested within `continent`); on the other hand, "crossing" means that a variable is crossed with another when every category of one variable co-occurs with every category of the other (for example, `country` crosses with `gender`). Nesting is a special case of crossing. The former exhibits a hierarchical ordering with lower-level category (children) nested within higher-level category (parent) (more categories in lower level); whereas the ordering of the latter makes no difference. Whilst constructing tidy temporal data, users also need to specify the key that identifies the structure on the data. Rather than creating a separate object that contains the data structure, like what the **hts** does, we propose a syntax-based interface to directly deal with variables, so that it takes advantage of data semantics. The expression of a vertical bar (`|`) is used to indicate nesting variables, while a comma (`,`) for crossing variables. @wilkinson2006grammar discussed the distinction between nesting and crossing with respect to facets in graphics, and he used `/` and `*` instead for nested and crossed expressions. A crossing structure displays all unique and possible combinations of crossed variables, including those not found in the data, in a graphical layout. However, tidy temporal data finds only the combinations that occur in the data and intentionally respects the structure of incomplete combinations.

@@

This "tidy" data representation choice will prove suitable for almost all conceivable applications of temporal data. But the data abstraction cannot live alone with an additional toolchain in order to be able to acquire insight and knowledge about the data itself. A desirable toolbox chains transformation, visualisation, and modelling as a data science pipeline. Each component itself is also a self-contained pipeline that can be decomposed into many individual operations. This paper will focus on part of the general pipeline---data transformation, for example filtering observations, selecting, and summarising variables. We shall see how this data abstraction breeds a consistent data process.

## Expressive verbs

@@

@@

# Application: U.S.A domestic flights on-time performance (2016-2017)

@@

The ground of time series modelling or forecasting is left untouched in this paper. The future plan is to bridge the gap between tidy data and model building. Currently, it is required to casting to matrix from tidy data and therefore building a model. But time series models should be directly applied to tidy data as other wrangling tools do, without such an intermediate step. In particular, a univariate time series model, like arima and exponential smoothing, can be applied to multiple time series independently. A tidy format to represent model summaries and forecasting objects will be developed and implemented in the future. Model summaries include coefficients, fitted values, and residuals; forecasting objects include future time path and distributions generating prediction intervals.

>>>>

In the literature, time series and panel (longitudinal) data are common terms referred to temporal-context data, depending on the research fields. Researchers who are concerned with modelling large $T$ and small $N$ would name as "time series" (serial correlation); those who are interested in modelling small $T$ and large $N$ as "panel data" (asymptotic). The data format is two-dimensional array, but different modelling focuses lend the data input to different representations. A matrix is used to represent multivariate time series where each row represents observations measured at a time point and each column represents a series ("wide" form). This matrix representation requires homogeneity (that is, all the columns must be of same type.), time indices implicitly inferred as attributes or meta-information, series of same time length, and explicit missingness. By contrast, panel data are organised in rectangular form of heterogeneous column types where multiple study subjects are stacked and repeated for its time indices in a single column ("long" form), due to commonly arisen unbalanced panels. This specification requires explicitly declared panel variable and index, which has been implemented in Stata's time series module and R package **plm** [@R-plm]. Evidently, this data organisation appears more flexible than matrix in supporting explicit time index, multiple subjects of different time lengths, and implicit missing values. less restrictions to be made. [R]

@@

The choice of representation of temporal-context data is made from a data-centric perspective, which is taken in the light of the operations that are to be performed on the data. This data abstraction reflects most common problems of incoming data and transformation in reality. Firstly, a data set must be structured in a "tidy" rectangular layout each row has the same schema and each field has a single value. Secondly, declaring the data set to contain temporal observations is determined by an "index" that represents time and a "key" that uniquely identifies each unit that measurements take place on over time. The "key" serves a similar purpose as the panel variable in the Stata's **tsset** command to define the units or subjects, but it is expanded to include multiple variables rather than a single one. A syntax is introduced to express a key consisting of nested and crossed factors. The composition of index and key uniquely defines each observation in a data table, which is equivalent to a primary key [@codd_relational_1970] in a relational database.

Table \@ref(tab:tb-sub) shows a "tidy" temporal data example using a subset of tuberculosis cases estimated by @tb-data. It contains 12 observations and 5 variables arranged in a "long" tabular form. Each observation uniquely records the number of people, who are diagnosed tuberculosis for each gender at three selected countries in the years of 2011 and 2012, where column `year` is declared as the index variable, columns `country`, `continent` and `gender` as key. To be more specific, the key is passed via `id(country | continent, gender)`, which produces a geographic hierarchy of country nested within continent, crossed with a demographic factor. Detailed discussions about index and key specifications will be given in Section \@ref(sec:index) and \@ref(sec:key). Although column `count` is the only measure in this case, it is sufficiently flexible to hold other measured variables, for example, adding the corresponding population size (if known) in order to calibrate the count later.

Given the nature of temporal ordering, a temporal data set must be sorted by time index as a result. If a key is explicitly declared, the key will be sorted first depending on the ordering of factor levels and the ordering of variables, and followed by arranging time in past-to-future order. The tuberculosis example shows the key is sorted in alphabetic order, followed by ascending years by default, as in Table \@ref(tab:tb-sub). Since time-based operations, for example leading, lagging, or differencing series, require a vector in its time order, the arranged data format assures its validity for these operations without returning unexpected results. There are other possible arrangements in the key, but the baseline is the index (i.e. `year`) ascends within each key element. If any operations on the tidy temporal data results in a deviation from this baseline, it will issue a warning to the users.

@@

The "key" identifies units or subjects that are observed over time in a data table, which are typically known in advance by users. The absence of key only occurs to a univariate case where the key is implicit. Multiple units must associate with an explicit "key" to allow for identification. The key is not constrained to a single column, but is comprised of multiple columns. When involving multiple variables, one need to distinguish nesting and crossing data variables. "Nesting" refers to a scenario where a variable is nested within another when each category of the former variable co-occurs with only one category of the latter (for example, `country` nested within `continent`); on the other hand, "crossing" means that a variable is crossed with another when every category of one variable co-occurs with every category of the other (for example, `country` crosses with `gender`). Nesting is a special case of crossing. The former exhibits a hierarchical ordering with lower-level category (children) nested within higher-level category (parent) (more categories in lower level); whereas the ordering of the latter makes no difference. Whilst constructing tidy temporal data, users also need to specify the key that identifies the structure on the data. Rather than creating a separate object that contains the data structure, like what the **hts** does, we propose a syntax-based interface to directly deal with variables, so that it takes advantage of data semantics. The expression of a vertical bar (`|`) (conditional) is used to indicate nesting variables, while a comma (`,`) for crossing variables. @wilkinson2006grammar discussed the distinction between nesting and crossing with respect to facets in graphics, and he used `/` and `*` instead for nested and crossed expressions. A crossing structure displays all unique and possible combinations of crossed variables, including those not found in the data, in a graphical layout. However, tidy temporal data finds only the combinations that occur in the data and intentionally respects the structure of incomplete combinations.

@@

This "tidy" data representation choice will prove suitable for almost all conceivable applications of temporal data. But the data abstraction cannot live alone with an additional toolchain in order to be able to acquire insight and knowledge about the data itself. A desirable toolbox chains transformation, visualisation, and modelling as a data science pipeline. Each component itself is also a self-contained pipeline that can be decomposed into many individual operations. This paper will focus on part of the general pipeline---data transformation, for example filtering observations, selecting, and summarising variables, by contrast to data cleaning. We shall see how this data abstraction breeds a consistent data process.

All the individual operations strive to keep the time context so that they will implicitly preserve the index variable.

## Modularity and clarity

@@

## Composition and transparency

@@

# Case studies

## U.S.A domestic flights on-time performance (2016-2017)

@@

A new visualization framework for handling time series with nesting and crossing structure is under development. The proposed data structure proves a natural input that can be piped into a visualization program. As long as the program understands the embedded structure, it will produce sensible plots.

The ground of time series modelling or forecasting is left untouched in this paper. The future plan is to bridge the gap between tidy data and model building. Currently, it is required to casting to matrix from tidy data and therefore building a model. But time series models should be directly applied to tidy data as other wrangling tools do, without such an intermediate step. In particular, a univariate time series model, like ARIMA and Exponential Smoothing, can be applied to multiple time series independently. A tidy format to represent model summaries and forecasting objects will be developed and implemented in the future. Model summaries include coefficients, fitted values, and residuals; forecasting objects include future time path and distributions generating prediction intervals.
